\documentclass{ximera}
\title{A Ferris Wheel}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Relating small changes as you ride ferris wheel.
\end{abstract}
\maketitle




\begin{question}  \label{Qererfeft4tr4}

This question is about a ferris wheel. The idea is this. After riding a ferris wheel for some time, the wheel gets stuck. Then it turns a through a small angle and gets stuck again. The question is to relate the small change in your height to the small change in the distance you move along your circular path as the wheel turns through this small angle.


\begin{enumerate}
\item The relationship between these small changes depends upon your position on the ferris wheel. The worksheet below illustrates this. Drag the slider $m$ in Line 2 to see how the relationship between the small change $\Delta h$ in your height and the small change in the distance $\Delta s$ along your path changes as your position changes.

\begin{onlineOnly}
    \begin{center}
\desmos{vc1xpi5mbq}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/vc1xpi5mbq}{151: Ferris Wheel 57}

Summarize your observations.
\begin{freeResponse}
\end{freeResponse}

\item Now let's focus on the particular position shown below at point $P$.

\begin{onlineOnly}
    \begin{center}
\desmos{9hlj4pkhni}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/9hlj4pkhni}{151: Ferris Wheel 58}

Drag the slider $n$ in Line 3 to see how the ratio $\Delta h/\Delta s$ changes as $\Delta s \to 0$. 

Then drag the slider $k$ in Line 4 and again the slider $n$ in Line 3 to make $\Delta s$ closer to $0$. Summarize your observations about how the ratio $\Delta h/\Delta s$ changes as $\Delta s \to 0$. 

\begin{freeResponse}
\end{freeResponse}


\item The first step in relating the small changes $\Delta h$ and $\Delta s$ is to find a function
\[
      h = f(s) \, , s\geq 0,
\]
that expresses your height (in feet) above the ground in terms of the distance (in feet) along your path since you boarded the wheel. 

For this, we'll assume the wheel has a radius of $30$ feet and that its center is $40$ feet above the ground.

\begin{onlineOnly}
    \begin{center}
\desmos{q6wnvqhnpg}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/q6wnvqhnpg}{151: Ferris Wheel 4}

Use the figure above to find an expression for the function $f$. Input your expression in the box below to check your work.
\[
    h  = f(s)  = \answer{40-30\cos(s/30)} .
\]

\item Now suppose the wheel stops after you have travelled just $15$ feet along your path. The wheel then turns through a small angle and stops again. Our aim is to determine if the small change
\[
       \Delta h = f(s) - f(15)
\] 
in your height is approximately proportional to the small change
\[
  \Delta s = s - 15
\]
in the distance you move along your path.

\begin{onlineOnly}
    \begin{center}
\desmos{uolmkpi5yr}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/uolmkpi5yr}{151: Ferris Wheel 3}

\begin{enumerate}
\item First find a function 
\[
   r = m(v)
\]
that expresses the average rate of change in your height with respect to your distance travelled between distances $s=15$ feet and $s=v$ feet. Enter that expression below with the appropriate domain.
\[
  r = m(v) = \answer{\frac{30\cos(1/2) - 30\cos(v/30)}{v-15}} \, , \,  v \geq 0 \text{ and } v\neq \answer{15}.
\]

\item Enter your expression for the function $r=m(v)$ in Line 2 of the worksheet below. Use the table and the slider $v$ in Line 4 to approximate the scaling factor that converts a small change 
\[
    \Delta s = s-15
\]
in your distance into an approximation of the change 
\[
  \Delta h = f(s) - f(15)
\] 
in your height. 


\begin{onlineOnly}
    \begin{center}
\desmos{cqprsr778n}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/cqprsr778n}{151: Ferris Wheel Table}

\item What are the units of the scaling factor?

\item What is the approximate relation between $\Delta h$ and $\Delta s$ for small changes $\Delta s$? Enter the scaling factor below, correct to 4 decimal places.

\[
      \Delta h \sim \answer{0.4794}\Delta s \text{ if } \Delta s \sim 0.
\]

\item Approximate the change in your height if you move an additional $0.04$ feet along your circular path from your position $s=15$ feet.

\end{enumerate}

\end{enumerate}


\end{question}

\end{document}
