\documentclass{ximera}
\title{Disco Dancing, an Optimization Problem}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Revolving light in a dance hall.
\end{abstract}
\maketitle

\begin{question}  \label{QdsadfKGDCCCC}
A spotlight in a circular dance hall of radius $r$ meters is located $b$ meters from the center of the hall. 

The light rotates at the constant rate of $\omega$ rad/sec. 

\begin{onlineOnly}
   \begin{center}
\desmos{vtzsgtxswm}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/vtzsgtxswm}{151: Disco Dancing 25}


\begin{enumerate}
\item Find an expression for the speed of the light beam as it moves along the wall at this instant. Your expression should be in terms of $\omega$, $r$, $b$. \emph{Hint:} Use the law of sines in $\Delta LOB$ to first find an expression for the measure of angle $\angle OLB$.

\item Check that your expression in  part (a) has the correct units.

\item Activate the Distance Function folder in Line 27 of the worksheet above. Then try to guess the position of the spotlight on the wall when its speed is increasing at the greatest rate. 

\item Determine the position of the spotlight on the wall when its speed is increasing at the greatest rate.

\end{enumerate}

\end{question}

%\begin{onlineOnly}
  % \begin{center}
%\desmos{vtzsgtxswm}{900}{600}
%\end{center}
%\end{onlineOnly}

%\href{https://www.desmos.com/calculator/vtzsgtxswm}{151: Disco Dancing 25}  %sp93wegrcb




\end{document}

