\documentclass{ximera}
\title{The Ladder and the Tree}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Implicit differentiation.
\end{abstract}
\maketitle

\begin{question} \label{Q43hhbnhhnnzzz}

A tree leans precariously with its trunk making an angle of $\phi = \pi/3$ radians with the ground. One end of a ladder leans against the trunk, the other rests on the horizontal ground. We analyze how a small change in the distance between the bottom of the ladder and the base of the trunk changes the distance between the top of the ladder and the base of the trunk.


\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/yojvjynme4}.
}
 
\begin{onlineOnly}
   \begin{center}
\desmos{yzhhdfdnw5}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/yzhhdfdnw5}{151: Ladder and Tree 20}

We'll let $u$ be the distance $OT$ between the top of the ladder and the base of the trunk (measured in feet) and  $s$ the distance $OB$ between the bottom of the ladder and the base of the trunk (also measured in feet).



We'll focus on the particular position of the ladder when the bottom end is $5$ feet from the trunk's base and the top end is $8$ feet \emph from the base \emph{as illustrated above}.

\begin{enumerate}

\item Use the slider $s$ approximate the derivative
\[
\frac{ds}{du}\Big|_{(s,u) = (5,8)} .
\]
Be sure to include units.

\item Find the length of the ladder

\item Find an equation that relates $s$ and $u$. Then activate the Folder in Line 20. Explain the significance of the curve and the slope of the tangent line to the curve at $P$.

\item Find an expression for the derivative $ds/du$.

\item Use your expression for $ds/du$ to evaluate the derivative
\[
\frac{ds}{du}\Big|_{(s,u) = (5,8)} .
\]

\item Interpret the meaning of the derivative using the language of small changes.

\item Use the slider $s$ to approximate the distance $s=OB$ when moving the top of the ladder a small distance away from the base of the tree makes the bottom end move twice as far toward the base. Then use calculus and algebra to apprximate that distance.

\item Activate the folder in Line 31 of the worksheet above.  Let $\theta$ be the angle (measured in radians) the ladder makes with the ground as marked. Take $\theta=2\pi/3$ when the ladder lies along the tree.

\begin{enumerate}
\item Use the slider $s$ to approximate the derivative 
\[
\frac{ds}{d\theta}\Big|_{s=5} .
\]
Include units.

\item Write an equation relating $\theta$ and $s$.

\item Find an expression for the derivative $ds/d\theta$ and use it to compute the exact value of the derivative
\[
\frac{ds}{d\theta}\Big|_{s=5} .
\]

\item Interpret the meaning of the above derivative using the language of small changes.

\item Find the exact value of the derivative
\[
\frac{du}{d\theta}\Big|_{s=5} .
\]

\end{enumerate}



\end{enumerate}

\end{question}


\end{document}