\documentclass{ximera}
\title{Stereographic Projection}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Maps of the earth, stereographic projection.
\end{abstract}
\maketitle


\section{Introduction to Differential Calculus}

\emph{Differential} has the same root as difference and this class, \emph{differential calculus}, is about calculating differences. Put another way, it is really about subtraction.

More specifically, the main question of differential calculus is how to relate changes in the output of a function to small changes in its inputs. For \emph{differentiable} functions, we can accurately approximate the change in the output by scaling a small change in the input. The \emph{scaling factor} is the \emph{derivative}.

You've seen scaling factors before. Look at a map of Seattle, for example, and you'll see a scale, maybe 3 miles/inch. But for a larger map, like one of the earth, there is no global scaling factor like this. Instead, the scaling factor varies from place to place and is usually a function of latitude. Even at a specific location, there is often not one, but infinitely many scaling factors, each depending on direction. 

But for a \emph{conformal} map (from the Latin \emph{conformalis}, \emph{of the same shape}) the scaling factor does not depend on direction. These maps preserve angles. The most well-know is probably the Mercator map. But we'll start with another.

\section{Stereographic Projection}

The worksheet below shows a stereogrphic projection of the earth between latitudes $-\pi/6$ and $\pi/2$, projected from the south pole.

\begin{onlineOnly}
    \begin{center}
\desmos{wy9zoftmbe}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/wy9zoftmbe}{151: Stereographic Projection World}


The worksheet here shows how the map is created. With the south pole $S$A at the top of the sphercical earth, a point $P$ on the surface of the earth gets mapped to the point $P$ in the plane of the equator by the line $\overrightarrow{SP}$. 

\begin{enumerate}
\item Drag the sliders $\theta_1$ (the longitude of $P$) and $\phi_1$ (the latitude of $P$) to see how the projection varies with latitude and longitude. Summarize your observations here.
\begin{freeResponse}
\end{freeResponse}
\end{enumerate}


\begin{onlineOnly}
    \begin{center}
\desmosThreeD{lnhm28tzak}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/lnhm28tzak}{152: Stereographic Projection}


\begin{onlineOnly}
    \begin{center}
\desmos{sdiaytbpn3}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/sdiaytbpn3}{151: Stereo Proj 2}



\begin{onlineOnly}
    \begin{center}
\desmosThreeD{dukkzdikzg}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/dukkzdikzg}{152: Stereographic Projection 2}










\end{document}