\documentclass{ximera}
\title{Practice with Limits and Derivatives}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Practice with limits and derivatives.
\end{abstract}
\maketitle


\section{More Practice with Rational Functions}

\begin{question} \label{QPODFDgdfgfg}
The function
\[
        W = f(r) = \frac{2000}{r^2} , r\geq 4 ,
\]
expresses the weight of an astronaut (measured in pounds) in terms of her distance from the center of the earth (measured in thousands of miles).

\begin{enumerate}
\item  Find a fully-simplified expression for the average rate of change in the astronaut's weight with respect to her distance from the earth's center between distances $r=b$ and $r=c$ thousands of miles from the center. Assume $b,c\geq 4$ and that $b\neq c$.

\item Use your \emph{expression from part (a) directly and the definition of the derivative} to find an expression (fully simplified) for the derivative
\[
   \frac{dW}{dr}\Big|_{r=b} .
\]

\item Use part (b) to evaluate the derivative
\[
   \frac{dW}{dr}\Big|_{r=5} .
\]

\item What are the units of the derivative above?

\item Explain the meaning of the derivative in part (c) using the language of scaling factors and small changes.
\end{enumerate}

\begin{explanation}
\begin{enumerate}
\item The average rate of change is
\begin{align*}
  \frac{\Delta W}{\Delta r} &= \frac{f(c) - f(b)}{c-b}  \\
                                       &= \left(\frac{1}{c-b}\right) \left( \frac{2000}{c^2} - \frac{2000}{b^2} \right) \\
                                       &=\left(\frac{2000}{c-b}\right) \left( \frac{b^2-c^2}{c^2 b^2} \right) \\
                                        &=\left(\frac{2000}{c-b}\right) \cdot \frac{(b+c)(b-c)}{c^2 b^2}  \\
                                        &= \frac{-2000(b+c)}{c^2 b^2} .
\end{align*}

\item The derivative is
\begin{align*}
           \frac{dW}{dr}\Big|_{r=b} &= \lim_{\Delta r\to 0} \frac{\Delta W}{\Delta r}  \\
                                                 &= \lim_{\Delta c\to b} \frac{-2000(b+c)}{c^2 b^2}  \\
                                                 &= -\frac{2000(b+b)}{b^2 b^2} \\
                                                 &= -\frac{4000}{b^3} .
\end{align*}

\item The value of the derivative is
\begin{align*} 
                \frac{dW}{dr}\Big|_{r=2}  &=  \left(-\frac{4000}{r^3}\right) \Big|_{r=5} \\
                                      &= -32 .
\end{align*}

\item The units of the derivative are pounds/(thousands of miles).

\item The derivative
\[
       \frac{dW}{dr}\Big|_{r=2} = -32 \,\,  \frac{\text{lbs}}{\text{thousands of miles}}.
\]
is a scaling factor that converts a small change
\[
 \Delta r = r-5
\]
in the astronaut's distance (in thousands of miles) from the center of the earth into an approximation for the small change
\[
   \Delta W = W = f(5) = W - 80
\]
in her weight. This means that for $\Delta r \sim 0$
\[
   \Delta W \sim -32\Delta r .
\]
So if the astronaut's distance to the earth's center increased from $r=5$ miles to $r=5.1$ miles, her weight would change by approximately
\begin{align*}
  \Delta W  &\sim \left(  -32 \, \, \frac{\text{lbs}}{\text{thousands of miles}} \right) \left( 0.1 \text{ thousands of miles}  \right) \\
                &= -3.2 \text{ lbs} .
\end{align*}

\end{enumerate}
\end{explanation}
\end{question}


\begin{question}  \label{Qldfer33}
\begin{enumerate}
\item Find a function
\[
   G = f(P) \, , \, P>0 ,
\]
that expresses the number of gallons of gas you can buy with $\$100$ in terms of the price (measured in dollars/gallon).

\item Sketch a graph of the function $f$ by hand.

\item Find a fully-simplified expression for the average rate of change in the number of gallons of gas you can buy with $\$100$ with respect to the price (measured in dollars/gallon) between prices $P=a$ and $P=b$, both measured in dollars/gallon.

\item Use your \emph{expression from part (c) directly and the definition of the derivative} to find an expression (fully simplified) for the derivative
\[
   \frac{dG}{dP}\Big|_{P=a} .
\]

\item What are the units of the derivative? Do \emph{not} simplify the units.

\item Evaluate the derivative
\[
  \frac{dG}{dP}\Big|_{P=5} .
\]

\item  Explain the meaning of the derivative in part (f) using the language of scaling factors and small changes.
\end{enumerate}
\end{question}


\begin{question} \label{Q9df93rv}
Between speeds of $20$ miles/hour and $70$ miles/hour the gas mileage of a car (in miles/gallon) is a quadratic function $G=f(v)$ of its speed (in miles/hr). The graph of the function $f$ is shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{6bbnti8zm5}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/6bbnti8zm5}{151: Gas Mileage Quadratic Function}

\begin{enumerate}
\item Find an expression for the function $G=f(v)$.

\item Find a fully-simplified expression for the average rate of change in the gas mileage (in miles/gal) with respect to speed (measured in miles/gallon) between speeds $v=a$ and $v=b$, both measured in miles/hour.

\item  Use your \emph{expression from part (b) directly and the definition of the derivative} to find an expression (fully simplified) for the derivative
\[
   \frac{dG}{dv}\Big|_{v=a} .
\]

\item What are the units of the derivative? Do \emph{not} simplify the units.

\item Evaluate the derivative
\[
  \frac{dG}{dv}\Big|_{v=30} .
\]

\item  Explain the meaning of the derivative in part (e) using the language of scaling factors and small changes.

\end{enumerate}
\end{question}



\begin{question} \label{QOidIdtgt4n}
The function
\[
   g = f(v) = \frac{1}{2}v + 10 \, , \, 10\leq v \leq 55 ,
\]
expresses the gas mileage (in miles/gal) of a car in terms of its speed.

\begin{enumerate}
\item At what rate (in gal/hr) does the car burn gas at a speed of $40$ miles/hour?

\item Find a function
\[
    r =g(v)\, , \, 10\leq v \leq 55 ,
\]
that expresses the rate (in gal/hr) at which the car burns gas in terms of its speed.

Answer: The function is 
\[
     r = g(v) = \answer{\frac{v}{\frac{1}{2}v + 15}} \, , \, 10\leq v \leq 55 .
\]

\item Find an expression for the average rate of change in the rate at which the car burns gas with respect to its speed between speeds of $v=w$ and $v=b$ miles/hour. Then use this expression and the algebra of limits to find an expression for the derivative
\[
   \frac{dr}{dv}\Big|_{v=b} .
\]

\item Evaluate the derivative
\[
   \frac{dr}{dv}\Big|_{v=40} .
\]

\item What are the units of the derivative above.? Do not simplify the units.

\item Explain the meaning of the derivative in part (d) using the language of scaling factors and small changes.


\end{enumerate}
\end{question}

\section{The Squaring Function and Its Inverse}

\begin{question}  \label{QPodfoitte43}
The function
\[
         A = f(s) =s^2 \, , \, s\geq 0 ,
\]
expresses the area of a square (measured in square feet) in terms of its side length (in feet).

\begin{enumerate}
\item Find an expression for the average rate of change in the area of a square with respect to its side length between side lengths $s=b$ and $s=w$ feet.

\item Use your expression from part (a) and the algebra of limits to find an expression for the derivative
\[
      \frac{dA}{ds}\Big|_{s=b}.
\]

\item Use the result of part (b) to evaluate the derivative
\[
   \frac{dA}{ds}\Big|_{s=5}.
\]

\item What are the units of the derivative in part (c).

\item Explain the meaning of the derivative (c) in terms scaling factors and small changes.

\begin{explanation}
You should have found that
\[
    \frac{dA}{ds}\Big|_{s=5} = 10 \text{ ft}^2/\text{ft} .
\]
This is a scaling factor that converts a small change
\[
    \Delta s = s-5 \sim 0,
\]
in the side length to a good approximation of the change  
\[
   \Delta A = f(s) - f(5) = f(s)-25
\]
in the area. That is, for $\Delta s \sim 0$,
\[
  \frac{\Delta A}{\Delta s} \sim \frac{dA}{ds}\Big|_{s=5} = 10 \text{ ft}^2/\text{ft},
\]
and
\[
   \Delta A \sim 10 \Delta s. 
\]

For example, if 
\[
 \Delta s = 0.1 ,
\]
then 
\[
    \Delta A \sim (10 \text{ ft} ) (0.1 \text{ ft}) = 1 \text{ ft}^2.
\]
Compare this approximation with the actual change
\[
   \Delta A = f(5.1) - f(5) = 1.01 \text{ ft}^2
\]
in the area. Pretty good.
\end{explanation}

\end{enumerate}

\end{question}

\begin{question} \label{QDfg5rtghbcvdf}
The function 
\[
   s = g(A) =\sqrt{A} \, , \, A\geq 0 ,
\]
expresses the side length (in feet) of a square in terms of its area (in square feet).

\begin{enumerate}
%\item Find an expression for the average rate of change in the side length with respect to area between areas $A=a$ and $A=w$ square feet.

\item Use the algebra of limits to find an expression for the derivative
\[
      \frac{ds}{dA}\Big|_{A=a}.
\]

\item Evaluate the derivative
\[
     \frac{ds}{dA}\Big|_{A=25}.
\]
What are its units? Explain its meaning in terms of small changes.

\begin{explanation}
The standard approach goes something like this.
\begin{align*}
    \frac{ds}{dA}\Big|_{A=a} &= \lim_{w\to a} \frac{g(w)-g(a)}{w-a} \\
                                             &= \lim_{w\to a} \frac{\sqrt{w}-\sqrt{a}}{w-a} \\
                                            &= \lim_{w\to a} \frac{\sqrt{w}-\sqrt{a}}{(\sqrt{w}-\sqrt{a})(\sqrt{w}+\sqrt{a})} \\
                                           &= \lim_{w\to a} \frac{1}{\sqrt{w}+\sqrt{a}} \\
                                            &= \frac{1}{2\sqrt{a}} .
\end{align*}

This approach obscures a key point. Namely, that the function $g$ is the \emph{inverse} of the function $f$ in Question 1. We can exploit this by instead rewriting the equation
\[
       s = \sqrt{A} = g(A)
\]
as 
\[
    A = s^2 = f(s).
\]
Now we'll find an expression for the derivative
\[
     \frac{ds}{dA}\Big|_{s=b}
\]
in terms of the side length $s=b$. This gives
\begin{align*}
        \frac{ds}{dA}\Big|_{s=b} &= \lim_{w\to b}\frac{w-b}{f(w) - f(b)} \\
                                              &=\lim_{w\to b}\frac{w-b}{w^2 - b^2)}\\
                                              &=\lim_{w\to b}\frac{1}{w+b)} \\
                                               &= \frac{1}{2b} .
\end{align*}

Letting $a = b^2$ be the area, we have $b=\sqrt{a}$, and
\[
  \frac{ds}{dA}\Big|_{A=a} = \frac{1}{2\sqrt{A}}.
\]

This makes it clear that the derivative of the inverse of a function is the \emph{reciprocal} of the function's derivative.

\end{explanation}


\item Use the result of part (b) to evaluate the derivative
\[
   \frac{dA}{ds}\Big|_{s=5}.
\]

\item What are the units of the derivative in part (c).

\item Explain the meaning of the derivative (c) in terms of small changes.


\end{enumerate}
\end{question}

\section{The Cubing Function and its Inverse}

\begin{question} \label{QRDFGGERE}
Repeat the two questions of the previous section with the function
\[
    V = f(s) = s^3 \, , \, s\geq 0 ,
\]
that expresses the volume of a cube in terms of its side length.
\end{question}


\section{A Square Root Function}
\begin{question} \label{QLDFLDFDl3reffgbf} 
A rock dropped from a height of $100$ feet falls to the surface of the planet Krypton without air resistance.

The function
\[
   v = f(h) = 6\sqrt{100-h} \, , \, 0\leq h \leq 100,
\]
expresses the rock's speed (in ft/sec) in terms of its height (in feet).

\begin{enumerate}
\item Find an expression for the average rate of change in the rock's speed with respect to its height between heights  $h=v$ and $h=b$ feet.

\item Use your \emph{expression from part (a) directly and the definition of the derivative} to find an expression (fully simplified) for the derivative
\[
   \frac{dv}{dh}\Big|_{h=b} .
\]

\item Use part (b) to evaluate the derivative
\[
   \frac{dv}{dh}\Big|_{h=36} .
\]

\item What are the units of the derivative above?

\item Explain the meaning of the derivative in part (c) using the language of small changes.
\end{enumerate} 


\end{question}

\section{More on the Squaring Function}

Check back later.


\end{document}