\documentclass{ximera}
\title{Short Quizzes Math 151}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Short Quizzes Math 151.
\end{abstract}
\maketitle

\begin{question}  \label{Qddsfsggtgttrghhgfgd}
The function
\[
      h = f(v) \, , \, 80 \leq v \leq 120 ,
\]
expresses the height of a helicopter (measured in feet) in terms of its speed (measured in ft/sec).

Suppose
\[
   \frac{dh}{dv}\Big|_{v=100} = - 20 .
\] 

\begin{enumerate}
\item What are the units of the above derivative?

\item Explain the meaning of the derivative using the language of small changes.
\end{enumerate}

\end{question}

\begin{question} \label{Qbgythjhjhgg}
The function
\[
     P = f(t) = 40e^{-t/20}\, , \, 0\leq t \leq 20 ,
\]
expresses the popuation (in millions) of a colony of bacteria in terms of the number of hours past noon.

\begin{enumerate}
\item Find a function
\[
     r  = g(P),
\]
that expresses the growth rate (in millions of bacteria/hour) of the population in terms of the population (in millions of bacteria).

\item Use the result of part (a) to determine the population when it is decreasing at the rate of $1,820,000$ bacteria/hour.

\begin{enumerate}
\item No credit for other methods.
\item End with a concluding sentence.
\end{enumerate}

\end{enumerate}
\end{question}

\begin{question}  \label{Qgghghnbnvxbvvb}
Find an equation of the tangent line to the curve
\[
   x^3\sin(5y) + 4e^{xy} = 4x
\]
at the point $(1,0)$.

\begin{enumerate}
\item Show all work.

\item Use the Leibniz notation for derivatives and their evaluations.

\item Include a brief explanation with your work.

\item End with a concluding sentence.
\end{enumerate}
\end{question} 

\end{document}