\documentclass{ximera}
\title{Implicit Differentiation}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
An introduction to implicit differentiation.
\end{abstract}
\maketitle

\section*{Discussion Questions}

\begin{question}  \label{Q:dsftr4tr656t}
Which of the following equations define $y$ implicty as a function of $x$ in a sufficiently small neighborhood of the given point? Supplement your reasoning with a graph of each relation.

(a) $x^2 + y^2 = 25$ near the point $(4,-3)$

(b) $x^2 + y^2 = 25$ near the point $(0,-5)$

(c) $x^2 + y^2 = 25$ near the point $(-5,0)$

(d) $x^2 - xy + y^2 = 1$ near the point $(1,1)$.

(e) $x^2 - xy + y^2 = 3$ near the point $(1,2)$.

(f) $x^2 - xy + y^2 = 3$ near the point $(2,1)$.

\end{question}

\section*{Introduction to Implicit Differentiation}

\begin{question}  \label{QDgbfb5566}
Let $p = f(t)$ be a differentiable function of $t$. Find expressions for each of the following derivatives, first supposing that
\[
       p = f(t) = t^3 + 1,
\]
and then more generally, not assuming any particular expression for the function $f$.

\pskip

(a) $\frac{d}{dp} \left(  p^5 \right)$

(b) $\frac{d}{dt} \left(  p^5 \right)$

(c)  $\frac{d}{dt} \left(  e^{2p} \right)$

(d) $\frac{d}{dt} \left( t^4 \sin(p) \right)$

\end{question}


\begin{question}  \label{Q:(dfdst43mmm}
(a) Find equations of the tangent lines to the ellipse
\[
       x^2 - xy + 2y^2 = 9
\] 
at the points where the ellipse intersects the $x$-axis.

(b) Find equations of all horizontal tangent lines to the ellipse.

(c) Find equations of all vertical tangent lines.

\begin{onlineOnly}
    \begin{center}
\desmos{ox3wvfitqk}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/ox3wvfitqk}{151: Implicit Ellipse}

\end{question}

\begin{question}  \label{Q:KKmbVC}
(a) Find an equation of the tangent line to the curve
\[
     4xe^y - 3y \cos(2xy)=0
\]
at the origin.

(b) Enter your equation in Line 5 of the worksheet below to check your answer.

\begin{onlineOnly}
    \begin{center}
\desmos{1qprxg226m}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/1qprxg226m}{151: Implicit 1}

\end{question}




\section*{Astroids}


\begin{question}  \label{QPDDFDFgg}
(a) The coordinate axes cut out a segment from the tangent line to the curve
\[
     x^{2/3} + y^{2/3}= 5
\]
at the point $P(8,1)$. Find the length of that segment. Do this by using implicit differentiation to help find an equation of the tangent line at $P$ and go from there. But first show that $P$ lies on the curve.

(b) As a bug crawls around the above curve and passes the point $P(8,1)$, it is moving away from the $x$-axis at the rate of $4$ cm/sec. Is the bug moving toward or away from the $y$-axis at this time? At what rate?  

(c) Find the slope of the tangent line in part (a) without using implicit differentiation.

%(c) Use geometry to find  the length of the segment cut out by the coordinate axes from the tangent line to the same curve 
%at the points where the curve intersects the line $y=x$.

%(d) Comments? Conjectures?


\end{question}


\begin{question}  \label{Q:LLKKKMM}
(a) Use implicit differentiation to show that segments cut by the coordinate axes from the tangent lines to the astroid
\[
   x^{2/3} + y^{2/3} = a^{2/3}
\]
all have the same length. Here $a>0$ is a constant.

(b) Prove the same result by using trigonometric functions to parameterize the astroid.

\begin{onlineOnly}
    \begin{center}
\desmos{vrythrvjuc}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/vrythrvjuc}{151: Astroid}


\end{question}



\section*{Ellipses and Related Curves}

\begin{question}  \label{Q:4bbfdellk}
(a) Use implicit differentiation to find an equation of the tangent line to the ellipse 
\[
   \frac{x^2}{a^2} + \frac{y^2}{b^2} = 1
\]
at the point $P$ with coordinates $(x_0, y_0)$. Here $a$, $b$ are positive constants.

(b) Input your equation from part (a) on Line 3 of the desmos worksheet below. Drag point $P$ to check your equation is correct.

(c) Solve part (c) again \emph{without} calculus by considering a composition of transformations that takes the circle
\[
       x^2 + y^2 = 1
\]
to the ellipse in part (a).

\begin{onlineOnly}
    \begin{center}
\desmos{t1e9v7ncpv}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/t1e9v7ncpv}{151: Tangents to Ellipse}

\end{question}


\begin{question} \label{Q:ergbbfrr}
An ellipse through the point $P(0,b)$ has focal points $F_1$ at the origin and $F_2$ at the point $A(a,0)$.

(a) Use the definition of an ellipse as the set of points whose distances to the foci have a constant sum to find an equation of the ellipse.

(b) Use implicit differentiation to find the slope of the tangent line to the ellipse at $P$.

(c) Find an equation of the line normal to the ellipse at $Q$.

(d) Find the coordinates of the point $Q$ where the normal line intersects the $x$-axis.

(e) Express the ratio $F_1 Q : Q F_2$ in terms of $a$ and $b$. Interpret the ratio geometrically.

The ratio is
\[
    \frac{F_1 Q}{Q F_2} = \answer{\frac{b}{\sqrt{a^2+b^2}}} . 
\]

(f) Check your work in the demonstration below.

\begin{onlineOnly}
    \begin{center}
\desmos{xqywotxxf9}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/xqywotxxf9}{151:Normal to an Ellipse}



\end{question}

\begin{question}  \label{Q:JDJHDHD}
Let $F_1$ and $F_2$ be respectively the origin and the point with coordinates $(a,0)$. The curve ${\cal C}$ passes through the point $P(0,b)$. The curve is defined by the property that the sum of the distance from a point $Q$ of ${\cal C}$ to $F_1$ and $k$ times its distance to $F_2$ is a constant.

(a) Find an equation of the curve ${\cal C}$.

(b) Find an equation of the tangent line to ${\cal C}$ at $P$.

(c) Find an equation of the line normal to ${\cal C}$ at $P$.

(d) Find the coordinates of the point $Q$ where the normal line intersects the $x$-axis.

(e) Express the ratio $F_1 Q : Q F_2$ in terms of $a$ and $b$. Interpret the ratio geometrically.

The ratio is
\[
    \frac{F_1 Q}{Q F_2} = \answer{\frac{bk}{\sqrt{a^2+b^2}}} . 
\]

(f) Check your work in the demonstration below.

\begin{onlineOnly}
    \begin{center}
\desmos{uz7w5uh0vl}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/uz7w5uh0vl}{151: Generalized Ellipse}

\end{question}


\section*{The Sliding Ladder}
\begin{question}  \label{Q:43543554LL}
You walk in along a straight path, moving either directly toward or directly away from a tree at the end of the path.

Let $s = f(t)$ be a function that expresses your distance to the tree in terms of the number of minutes past noon and suppose that
\[
     \frac{ds}{dt}\Big|_{t=5} = -240 .
\]

(a) Interpret the meaning of the above derivative.

(b) What is your speed at 12:05pm?

(c) Find an expression for your speed at time $t$ minutes past noon.
\end{question}


\begin{question}  \label{QMMnd6793}
A tree leans precariously with its trunk making an angle of $\phi$ radians with the ground. One end of a ladder with length $L$ feet leans against the trunk, the other rests on the horizontal ground. You slide the bottom end of the ladder along the ground at random.

(a) Let $s$ and $t$ be the respective distances (measured in feet) from the base of the trunk to the bottom and top ends of the ladder. Find an equation that relates these distances. Use the parameters $L$ and $\phi$, \emph{not} their particular values in the demonstration below.

(b) Let $u$ be time measured in seconds and suppose that $s$ and $t$ are (unknown) functions of $u$. Express the derivative $dt/du$ in terms of the derivative $ds/du$. Interpret the meanings of the these derivatives.

(c) Switch $s$ and $t$ in your equation from part (b). What do you notice? Explain why.

(d) Use part (b) to express the speed of the top end of the ladder in terms of the speed of its bottom end.

(e) Suppose $\phi=\pi/3$ and that at some instant the bottom end of the ladder is three times as far from the trunk's base as the top end. What can you say about the speeds of the ends of the ladder at this instant? 

(f) Still supposing $\phi=\pi/3$, find all possible angles the ladder makes with the ground when the top end of the ladder is moving four times as fast as the bottom end. Characterize each angle in one of two ways:

(i) One end of the ladder is moving toward the trunk's base and the other end away from the base.

(ii) Both ends of the ladder are either moving toward or way from the trunk's base.

Give exact angles. Then use a calculator to approximate their radian measures to the nearest hundredth.

(g) Use the animation below to check that your answers to parts (e) and (f) are reasonable. The graph is of the relation


\begin{onlineOnly}
    \begin{center}
\desmos{58jqjo0inl}{900}{600}
\end{center}
\end{onlineOnly}

Geogebra activity available at \href{https://www.desmos.com/calculator/58jqjo0inl}{151: Tree and Ladder}

\end{question}



\begin{question}  \label{Q:KNDMMD}
A tree leans precariously with its trunk making an angle of $\phi$ radians with the ground. One end of a ladder with length $L$ feet leans against the trunk, the other rests on the horizontal ground. You slide the bottom end of the ladder along the ground at random.

Suppose at some instant the ladder makes an angle of $\beta_0$ radians with the trunk and that the bottom end is $s_0$ feet from the trunk's base. Suppose also that the ladder is rotating at the rate of $\omega_0$ rad/sec at this same moment. 

(a) Find an expression for rate of change in the distance from the bottom end of the ladder to the base of the trunk at this time.  

(b) Find an expression for rate of change in the distance from the top end of the ladder to the base of the trunk at this time.

(c) Find expressions for the speeds of the ends of the ladder at this time.

\begin{hint}
Use the law of sines in $\Delta BOT$ in the demonstration from the previous question.
\end{hint} 


\end{question}


\begin{question}  \label{Q:er43455}
The function
\[
    h = f(\delta) = \frac{24}{\pi} \arccos(-\tan\phi \tan\delta) \, , \, -\pi/2+\phi < \delta < \pi/2-\phi ,
\]
expresses the number of hours of daylight per day at latitude $\phi$ in terms of the declination of the sun. The declination of the sun ($\delta$) is the angle the sun's rays make with the plane of the equator, taken to be positive between the spring and fall equinoxes in the northern hemisphere. The latitude $\phi$, $-\pi/2 \leq \phi \leq \pi/2$, is positive at points in the northern hemipshere.

\begin{onlineOnly}
    \begin{center}
\geogebra{vnhrutwu}{900}{600}
\end{center}
\end{onlineOnly}

Geogebra activity available at \href{https://www.geogebra.org/classic/vnhrutwu}{151: Declination of Sun 2}


(a) Use the graph of the function $f$ below (at latitude $\phi \sim 1.1$) to sketch a graph of the derivative
\[
   \frac{dh}{d\delta} = f^\prime(\delta) .
\]

(b) Suppose the latitude $\phi$ is held constant and find an expression for the derivative 
\[
   \frac{dh}{d\delta} = \frac{d}{d\delta} \left(  -\tan\phi \tan \delta    \right).
\]

(c) What are the units of the derivative in part (b)?

(d) Input your expression for the derivative in Line 4 of the worksheet below (follow the directions there). Then vary the slider $\phi$ to see how the function $f$ and its derivative vary with latitude. Summarize your observations.

(e) Find an expression for the derivative 
\[
   \frac{dh}{d\delta}\Big|_{\delta = 0} =  \frac{d}{d\delta} \left(  -\tan\phi \tan \delta    \right)\Big|_{\delta =0}
\]
in terms of the latitude $\phi$. 

(f) Evaluate the derivative in part (e) at a latitude of $\phi =\pi/4$. Interpret its meaning in terms of small changes.

\begin{onlineOnly}
    \begin{center}
\desmos{ifomatkcta}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/ifomatkcta}{151: Length of Day 1}

\pskip 

(g) Suppose now that each month has 30 days so that there are 360 days in one year. Suppose also that the declination of the sun varies sinusoidally as a function of time, that the maximum declination of $\delta = 23.5^\circ$ occurs on the summer solstice (say June 21st) and the minimum declination $\delta = -23.5^\circ$ occurs on the winter solstice (December 21st). 

Find an expression for a function
\[
     \delta = k(t) \, , \, t\geq 0,
\]
that gives the declination of the sun (measured in radians) in terms of the number of days since the spring equinox. 

(h) Find a function
\[
   r = g(\phi) \, , \, -\pi/2 < \phi < \pi ,
\]
that expresses the rate of change in the number of \emph{minutes} of daylight per day (measured in (minutes of daylight/day)/day)) on the spring equinox in terms of the latitude $\phi$. Input this function in Line 1 of the demonstration below.

(i) Use the result of part (f) to approximate the rate in (hours of daylight/day)/day) at which the number of daylight hours per day is changing at a latitude of $\phi = \pi/4$ radians on the spring equinox and on the fall equinox. 

(j) Evaluate the rates from part (i) in Fairbanks, Alaska, latitude $64.8^\circ$N.

\begin{onlineOnly}
    \begin{center}
\desmos{nf8n5uphhl}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/nf8n5uphhl}{151: Length of Day 2}



\end{question}


\section*{Waves}

\begin{question}   \label{Q:DFLDFDFGggg}
The function
\begin{equation}  \label{Eq:Wave1}
      y= f(x,t) = a \sin (kx - \omega t) , t\geq 0,
\end{equation}
describes a wave on a string. The functions expresses the displacement (in meters) of a point on the string in terms of the position $x$ (in meters) of the point and time $t$, measured in seconds since the motion began.

(a) Experiment with the sliders $k$, $\omega$, in the demonstration below, playing the slider $u$ ($u$ is just another name for $t$). Summarize your observations. In particular, be sure to turn off Line 1 to be better able to see the motion of the individual points of the string.

(b) What are the units of $k$ and $\omega$? How do you know?

(c) Find an expression for the wavelength $\lambda$ in terms of $k$, $\omega$.

(d) Find an expression for the period of oscillation $T$ in terms of $k$, $\omega$.
 
(e) Hold $y$ constant and differentiate each side of equation (\ref{Eq:Wave1}) with respect to $t$ to find an expression for the speed of the wave. Turn on the graph in Line 9. Explain the logic behind the computation. 

\begin{onlineOnly}
    \begin{center}
\desmos{9xmkig9hwi}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/9xmkig9hwi}{151: Traveling Wave 1}

\end{question}


\end{document}