\documentclass{ximera}
\title{Riemann Sums}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Riemann sums and summation notation.
\end{abstract}
\maketitle



\begin{question} \label{QPdfer34334}
The function
\[
   v = g(s) = 40 + \frac{s}{5} \, , \, 20\leq s \leq 100 ,
\]
expresses the speed (in miles/hr) of a car in terms of the trip odometer reading (measured in miles) during part of a 200-mile trip.

\begin{enumerate}
\item Use summation notation for a Riemann sum with $10$ equal intervals of distance to find an upper bound for the time it takes to travel the $80$ miles in this portion of the trip. Enter this in Line 1 of the worksheet below to get an approximation for the time. Activate the appropriate folder in Line 12 (Left endpoints) or Line 19 (Right Endpoints) to interpret the sum geometrically.

\item Use summation notation for a Riemann sum with $10$ equal intervals of distance to find a lower bound for the time it takes to travel the $80$ miles in this portion of the trip. Enter this in Line 1 of the worksheet below to get an  approximation for the time.
Activate the appropriate folder in Line 12 (Left endpoints) or Line 19 (Right Endpoints) to interpret the sum geometrically.

\item Use summation notation for a Riemann sum with $n$ equal intervals of distance to find an upper bound for the time it takes to travel the $80$ miles in this portion of the trip. Enter this in Line 1 below. Then drag the slider $n$ in Line 2 to get a more accurate upper bound.

\item Use summation notation for a Riemann sum with $n$ equal intervals of distance to find a lower bound for the time it takes to travel the $80$ miles in this portion of the trip. Enter this in Line 1 below. Then drag the slider $n$ in Line 2 to get a more accurate upper bound.


\end{enumerate}

\begin{onlineOnly}
    \begin{center}
\desmos{k10phbukon}{900}{600}
\end{center}
\end{onlineOnly}
 
\href{https://www.desmos.com/calculator/k10phbukon}{152: Distance Speed 2}
\end{question}

\begin{question} \label{QOErermern333}
The function
\[
 r = f(t) = 70 \cos^3 \sqrt{t+40} \, , \, 0\leq t \leq 48,
\]
expresses a balloon’s rate of ascent (in ft/min) in terms of the number of minutes past noon. The balloon was 300 feet high at 12:04pm.

\begin{enumerate}
\item Use summation notation for an expression with $10$ equal time intervals to find a lower bound for the balloon's height at 12:44pm. Enter this expression in Line 1 of the worksheet below to get an  approximation for the height. Activate the appropriate folder in Line 12 (Left endpoints) or Line 19 (Right Endpoints) to interpret the sum geometrically.

\item Use summation notation for an expression with $10$ equal time intervals to find an upper bound for the balloon's height at 12:44pm. Enter this expression in Line 1 of the worksheet below to get an  approximation for the  height. Activate the appropriate folder in Line 12 (Left endpoints) or Line 19 (Right Endpoints) to interpret the sum geometrically.


\item Use summation notation for an expression with $n$ equal time intervals to find a lower bound for the balloon's height at 12:44pm. Enter this expression in Line 1 of the worksheet below to get an  approximation for the height. Enter this in Line 1 below. Then drag the slider $n$ in Line 2 to get a more accurate upper bound.

\item Use summation notation for an expression with $n$ equal time intervals to find an upper bound for the balloon's height at 12:44pm. Enter this expression in Line 1 of the worksheet below to get an  approximation for the height. Enter this in Line 1 below. Then drag the slider $n$ in Line 2 to get a more accurate upper bound.

\begin{onlineOnly}
    \begin{center}
\desmos{qwuqprd9vf}{900}{600}
\end{center}
\end{onlineOnly}
 
\href{https://www.desmos.com/calculator/qwuqprd9vf}{152: Balloon 56}
 
\end{enumerate}
\end{question}


\begin{question}  \label{QIIIUER3355}
The function
\[
      \theta = g(s) = \frac{s}{2} - \frac{s^3}{12} \, , \, 0\leq s \leq 3 , 
\]
expresses the inclination angle (in radians) of the trail to Nada Lake in terms of the distance from the trailhead (measured along the trail in miles). The inclination angle is the angle the trail makes with the horizontal. It is positive (negative) when the trail slopes upward (downward) in the direction away from the trailhead.

The trail is at an elevation of $3200$ feet $0.5$ miles from the trailhead.

\begin{enumerate}
\item Use summation notation for an expression with $10$ equal intervals of distance to find a lower bound for the trail's elevation (in feet) at Nada Lake (3 miles from the trailhead). Enter this in Line 1 of the worksheet below to get an approximation for the elevation. Activate the appropriate folder in Line 12 (Left endpoints) or Line 19 (Right Endpoints) to interpret the sum geometrically.

\item Use summation notation for an expression with $10$ equal intervals of distance to find an upper bound for the trail's elevation (in feet) at Nada Lake. Enter this in Line 1 of the worksheet below to get an approximation for the elevation. Activate the appropriate folder in Line 12 (Left endpoints) or Line 19 (Right Endpoints) to interpret the sum geometrically.


\item  Use summation notation for an expression with $n$ equal intervals of distance to find an upper bound for the trail's elevation (in feet) at Nada Lake. Enter this in Line 1 below. Then drag the slider $n$ in Line 2 to get a more accurate upper bound.

\item Use summation notation for an expression with $n$ equal intervals of distance to find a lower bound for the trail's elevation (in feet) at Nada Lake. Enter this in Line 1 below. Then drag the slider $n$ in Line 2 to get a more accurate upper bound.

\begin{onlineOnly}
    \begin{center}
\desmos{qhyco8cuou}{900}{600}
\end{center}
\end{onlineOnly}
 
\href{https://www.desmos.com/calculator/qhyco8cuou}{152: Inclination Angle 34}

\end{enumerate}
\end{question}

\begin{question}  \label{Q9er34rre}
Repeat Question 3 with the function
\[
   m = g(s) = \frac{s}{2} - \frac{s^3}{12} \, , \, 0\leq s \leq 3 ,
\]
that expresses the slope of a trail in terms of the distance from the trailhead (measured along the trail in miles). The slope is measured with respect to the horizontal. It is positive (negative) when the trail slopes upward (downward) in the direction away from the trailhead.
\end{question}


\begin{question} \label{Q8fre}
The function
\[
      v = g(t) = 20+  40t - 3t^2 \, , \, 0\leq t\leq 1 ,
\]
expresses the speed of a car (in miles/hr) in terms of the number of hours past noon.

The function 
\[
    G = h(t) = 30 + 10t  \, , \, 0\leq t\leq 1
\]
expresses the gas mileage (in miles/gal) of that same car in terms of the number of hours past noon.

The car has $10$ gallons of gas in its tank at 12:15pm.

Answer the four parts of the previous questions in regard to approximating the number of gallons of gas in the tank at 1:00pm.


\end{question}



\end{document}
