\documentclass{ximera}
\title{Parametrically-defined Curves, CW}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Curves defined parametrically and signed area.
\end{abstract}
\maketitle

\section{The Basics}

Not all curves are the graphs of functions.

The green curve below in the first quadrant, for example, is neither the graph of $y$ as a function of $x$ (ie. $y=f(x)$) nor of $x$ as a function of $y$ (ie. $x=f(y)$).

Why not?
\begin{freeResponse}
\end{freeResponse}

\begin{onlineOnly}
    \begin{center}
\desmos{vio0b8ztnl}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/vio0b8ztnl}{152: Parametric 2}


But we can describe the curve \emph{parametrically}. This means to give functions
\[
      x = f(t) \text{ and } y=g(t)
\]
that express the coordinates $(x,y)$ of a point on the curve in terms of some parameter ($t$ here) which is often helpful to think of as being time. We can then think dynamically of the \emph{motion} of a point along its path instead of just a static path.

These coordinate functions for the green curve above are graphed in the fourth ($x=f(t)$) and second ($y=g(t)$) quadrants and share the domain $0\leq t\leq 4$ (you should ignore the negative signs on the vertical and horizontal axes). To see the motion of the point along its path, play the slider $u=t$ in Line 2. 

The question we would like to address here is to compute the unsigned (ie. positive) area of the region bounded by the curve 
\[
  (x,y) = (f(t), g(t)) \, , \, 0\leq t \leq 4 .
\]

We'll use with the demonstration below, where the graphs of the coordinate functions are hidden.

\begin{onlineOnly}
    \begin{center}
\desmos{z9ndh3wvvv}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/z9ndh3wvvv}{152: Parametric 2B}



As usual, our first choice is to decide which way to slice the region and we'll start by slicing perpendicular to the $x$-axis. But this is not really an accurate description because the differential slices run from the $x$-axis to the curve and do not always lie within the region. 

To see this, drag the slider the slider $u$ in Line 2 to move the differential rectangle. Sometimes the rectangle lies below the region, other times it lies partly within and partly below the region, and at other times completely within the region. Nevertheless, we can use these differential rectangles to compute the area of the region.

The key idea is to work with their \emph{signed} areas. Cutting perpendicular to the $x$-axis, the differential rectangles have signed area
\begin{align*}
 dA & = y \, dx \\
      & = g(t) d(f(t)) \\
      & = g(t) (f^\prime(t) \, dt) \\
      & = g(t) f^\prime (t) \, dt .
\end{align*}

And so the region bounded by the green curve and the $x$-axis has signed area
\[
   A = \int_0^4 g(t) f^\prime(t) \, dt.
\]

Let's stop and think a minute about what we mean by signed area. The sign of the differential area $dA$ depends upon \emph{both} the signs of $y=g(t)$ and $dx = f^\prime (t) \, dt$. For our example $y = g(t) \geq 0$, $dt>0$, and so $dA$ has the same sign as $dx$. 

\begin{enumerate}

\item Drag the slider $u$ in Line 2 from $u=0$ to $u=b=4$. For what values of $u$ is $dA$ positive? Negative?

\item Sketch by hand a rough graph of the cumulative signed area function
\[
      A_1(t) = \int_0^t g(u) f^\prime(u) \, du.
\] 
Then drag the slider $u$ in Line 2 to $u=0$ and activate the folder \emph{Graph of signed area function, slices perp to $x$-axis} in Line 7. Drag $u$ to $u=b=4$ and see the graph of the signed area function. Compare it with your graph.

\item How is the integral 
\[
  A_1(4) = \int_0^4 g(t) f^\prime(t) \, dt
\]
related to the area of the region bounded by the $x$-axi and the curve $(x,y) = (f(t), g(t))$? Explain your reasoning.
\begin{freeResponse}
\end{freeResponse}

We can also take our differential rectangles perpendicular to the $y$=axis. Then
\[
   dA = x \, dy = \answer{f(t) g^\prime(t)}\, dt .
\]

\item Repeat parts (a)-(c) above for these differential rectangles and the signed area function
\[
     A_2(t) = \int_0^t f(u) g^\prime(u) \, du .
\]

\item Now use the actual coordinate functions
\[
     x = f(t) = t^2 - 3t + 3 \, , \, 0\leq t \leq 4
\]
and
\[
   y = g(t) = 4t - t^2 \, , \, 0\leq t \leq 4
\]
to compute the area of the region bounded by the $x$-axis and the curve
\[
   (x,y) = (f(t), g(t)) \, , \, 0\leq t \leq 4.
\]
Do this twice. Once with slices perpendicular to the $x$-axis and again with slices perpendicular to the $y$-axis.

\end{enumerate}


\section{An Application: Gas Consumption}

\begin{example} \label{Ex:KDfdfer3dsferER}


The function
\[
    r = f(t) \, , \, 0 \leq t \leq 4,
\]
expresses the rate (in gal/mile) at which a car burns gas in terms of the number of hours past noon during a $240$-mile trip.

The function 
\[
      s = g(t) \, , \, 0\leq t \leq 4 ,
\]
expresses the car's trip odometer reading (in miles) in terms of the number of hours past noon.

The car has $14$ gallons of gas at 1:00pm.

The functions $f$ (second quadrant) and $g$ (fourth quadrant) along with the curve $(x,y)= (g(t),f(t))$ (first quadrant) are graphed below.

\begin{onlineOnly}
    \begin{center}
\desmos{akedzpal8z}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/akedzpal8z}{Substitution and Gas Comsumption 22}



\begin{enumerate}
\item Approximate the rate, measured in gal/hour, at which the car burns gas at 1:54pm. Explain your reasoning and include units for all numbers in your computation.

%\item Find an expression for the number of gallons of gas the car burns from 1:00pm until time $t$ hours past noon, $0\leq t \leq 4$. 

\item Approximate the number of gallons of gas in the tank at 1:54pm.

\item Find a function 
\[
      G = w(s) \, , \, 0 \leq t \leq 4,
\]
that expresses the number of gallons of gas in the car in terms of the number of hours past noon.


\end{enumerate}



\end{example}



\section{Exercises}

\begin{exercise}  \label{EOfidf3r3r2}

This question is about the circle
\[
     (x,y) = (f(\theta), g(\theta)) = (a \cos \theta , b+a\sin\theta) \, , \, 0\leq \theta \leq 2\pi .
\]
Here $a,b$ are positive constants measured in meters.

\begin{onlineOnly}
    \begin{center}
\desmos{weihqr1fxu}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/weihqr1fxu}{152: Parametric 3}

\begin{enumerate}

\item Use the worksheet below to sketch by hand the signed area functions
\[
     A_1(\theta) = \int_0^{t} g(t) f^\prime (t) \, dt
\]
and
\[
  A_2 (\theta) = \int_0^{t} f(t) g^\prime (t) \, dt .
\]
Then check your graphs by activating the folders in Lines 7 and 23.

\begin{onlineOnly}
    \begin{center}
\desmos{bg3tmflc1f}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/bg3tmflc1f}{152: Parametric 3B}


\item Evaluate the integrals
\[
       A_1(2\pi) = \int_0^{2\pi} g(\theta) f^\prime (\theta) \, d\theta
\]
and 
\[
   A_2(2\pi) = \int_0^{2\pi} f(\theta)g^\prime (\theta) \, d\theta 
\]
\emph{geometrically} by using the graph of the function
\[
     A = a^2 \sin^2\theta \, , \, 0\leq \theta \leq \pi
\]
shown below. Work in general, not with the specific value of $a$ in the worksheet.


\begin{onlineOnly}
    \begin{center}
\desmos{dsoqsyyfba}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/dsoqsyyfba}{152: Sine Squared}
\end{enumerate}
\end{exercise}

\begin{exercise} \label{EKdfe34dfds}
Use the parameteriziation
\[
    (x,y) = (a\cos\theta, a \sin\theta) \, , \, 0\leq \theta \leq 2\pi,
\]
of the circle with radius $a$ meters centered at the origin to find an expression for the volume of a sphere with the same radius.

\end{exercise}


\begin{exercise} \label{EKdfmdfee}

The circle with center $(x,z) = (b,0)$ and radius $a$ is rotated about the $z$-axis to sweep out the torus shown below.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{baapxxf50y}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/baapxxf50y}{152: Torus}

The aim of this problem is to express the volume of the torus in terms of the parameters $a$ and $b$, measured in meters. We'll do this by parameterizing the circle in terms of the polar angle of the vector $\overrightarrow{CP}$ from the center of the circle to a point $P$ on the circle with coordinates $(x,y)$. See the figure below.

\begin{onlineOnly}
    \begin{center}
\desmos{qjpoe06x3y}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/qjpoe06x3y}{152: Torus 2D}

\begin{enumerate}

\item Express the coordinates of point $P$ in terms of the polar angle $\theta$.

\item Activate the folder in Line 1 of the worksheet above to see the cross-section of the torus swept out by $PQ$. The cross-section should really be drawn peprendicular to the page. You can see the cross-section in the top worksheet by deactivating the folder \emph{Torus} in Line 4.

\item Express the area of the cross-section in terms of $a$, $b$, and $\theta$.

\item Express the differnential volume of the corresponding slice in terms of $a$, $b$, and $\theta$.

\item Add these differential volumes to find the volume of the torus.
\end{enumerate}
 


\end{exercise}



\begin{exercise} \label{Epdfsderenn}
Find the area bounded by the astroid
\[
 (x,y) = (a\cos^3\theta, a \sin^3\theta) \, , \, 0\leq \theta \leq 2\pi ,
\]
where $a>0$ is a constant measured in meters.
\end{exercise}


\begin{exercise}  \label{EKDferf3vbbb}
Find the area bounded by the ellipse
\[
   \frac{x^2}{a^2} + \frac{y^2}{b^2} = 1, 
\]
where $a,b>0$ are constants measured in meters. 

Start by parameterizing the ellipse.
\end{exercise}




\end{document}