\documentclass{ximera}
\title{Rates of Change, Gas Mileage}


\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Limits in context.
\end{abstract}
\maketitle


\pskip

\section{Gas Mileage} 

\begin{example} \label{Ex:9stghd8gfs}
The function 
\[
   G = f(s) = \frac{2}{5} +\frac{1}{5000} \left( 40(s+2)^2 - (s+2)^3 \right) , \, 0\leq s \leq 23 ,
\]
expresses the number of gallons of gas in your car in terms of your distance from home. The distance is measured in miles along your route. 

\begin{onlineOnly}
    \begin{center}
\desmos{xzkknfpkw3}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at
\href{https://www.desmos.com/calculator/xzkknfpkw3}{151: Gas as a Function of Distance}

\begin{enumerate}
\item Use the graph of the function $f$ shown above to determine if you are driving toward or away from home. Explain your reasoning.

\item Find your average gas mileage (in miles/gallon) over the interval $s \in [8,18]$.

\item Use the graph to approximate your gas mileage at the moment you are $18$ miles from home. Do this by zooming in on the appropriate point.

\item Find an expression for the function $r=h(s)$, that gives your average gas mileage between distances $s$ miles and $18$ miles from home. Include the correct domain.

\item Use the algebra of limits to determine your exact gas mileage at the moment you are $18$ miles from home.

%(d) Use the graph to approximate your distance from home when your car was getting the best mileage. Explain your reasoning. 

%(e) Use the graph to approximate your distance from home when your car was getting the least mileage. Explain your reasoning. 

%(f) Use the algebra of limts to determine your best and worst gas mileages during the trip.

\item Use the result of part (e) to approximate the change in the volume of gas
\[
    \Delta G = f(s)-f(18)
\]
in terms of the change
\[
   \Delta s = s - 18
\]
in your distance from home for values of $s$ near $18$ miles. What are the units of the scaling factor?

\item Use the result of part (f) to approximate your distance from home when there are $1.9$ gallons of gas in your tank.

%\item Find an expression for your average gas 

\end{enumerate}


\end{example}



\end{document}