\documentclass{ximera}
\title{On Trails}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Hiking bugs.
\end{abstract}
\maketitle

\section{Trails Desribed with Functions}

In this section we model a trail with a curve $h=f(x)$, where $x$ and $h$ are measured in meters. We take the $h$-coordinate to measure the height (in meters) above sea-level and suppose the positive $x$-axis points due east.

\begin{question}  \label{QOideIIEIEII}
A beetle crawls eastward along the trail $h = f(x)$ at a constant speed of $v_0$ meters/sec.

At what rate is the beetle gaining altitude if

\begin{enumerate}
\item $f(x) = x$ ?

\item $f(x)=2x$ ?

\item $f(x) = mx$ ? 

\item Find the limit of the beetle's rate of ascent as $m\to \infty$ as it crawls on the trail $h=mx$ at the constant speed of $v_0$ meters/sec.
\end{enumerate}
 
\end{question}

\begin{question}  \label{QPeriERE}
\begin{enumerate}

\item A beetle crawls along the parabola
\[
     h = 9-\frac{1}{4}\left(x-6\right)^{2} \, , \, 0\leq x \leq 12,
\]
(coordinates in meters) in the eastward direction at a constant speed of $v_0$ m/sec. At what rate is it ascending when it is $10$ meters above the ground for the second time?

\item A beetle crawls along the curve
\[
     h =4-3\cos\left(\frac{x}{2}\right) \, , \, 0\leq x \leq 12,
\]
(coordinates in meters) in the eastward direction at a constant speed of $v_0$ m/sec. At what rate is it ascending when it is $5$ meters above the ground for the second time?

\end{enumerate}

\end{question}




\section{Parametrically Defined Trails}


\end{document}