\documentclass{ximera}
\title{The Tangent Function and its Inverse}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
The tangent function and its inverse.
\end{abstract}
\maketitle


\section{The Tangent Function}

The most common way to find the derivative of the function $y=\tan(\theta)$ is to write 
\[
  \tan\theta = \frac{\sin\theta}{\cos \theta}
\]
and use the quotient rule. But this is not all that insightful, and we have not learned the quotient rule yet. Here are two other approaches.

\begin{question}  \label{QPDeredsRMNFR}

We'll use the figure below to find an expression for the derivative
\[
   \frac{d}{d\theta} \left( \tan\theta \right)
\]
as follows.

\begin{onlineOnly}
   \begin{center}
\desmos{10538a93e8}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/10538a93e8}{151: Tangent Derivative}  

\begin{enumerate}
\item First convince yourself that the marked angles $\theta = \angle BOA$ and $\angle QBC$ are congruent. Well almost, assuming $\Delta \theta \sim 0$.

\item Express the length $OB$ in terms of $\theta$.

\item Express the length $QB$ in terms of $\theta$ and $\Delta \theta$.

\item Express the length $BC$ in terms of $\theta$ and $\Delta \theta$.

\item Then find an expression for 
\[
   \frac{dy}{d\theta} = \frac{d}{d\theta} \left( \tan\theta \right) .
\]
\end{enumerate}
\end{question}

\begin{question} \label{Pfmf3901}
\begin{enumerate}
\item Find an expression for the average rate of change of the function
\[
      f(\theta) = \tan\theta
\]
between $\theta  =\theta_1$ and $\theta = \theta_2$.

\item Use part (a) and the algebra of limits to find an expression for the derivative
\[
     \frac{dy}{d\theta}\Big|_{\theta = \theta_1} =  \frac{d}{d\theta}\left( \tan\theta  \right)\Big|_{\theta = \theta_1} .
\]
\emph{Hint:} Use the identity
\[
      \sin(\alpha - \beta) = \sin \alpha\cos\beta - \cos\alpha \sin \beta .
\]

\item Repeat parts (a) and (b) for the function $f(\theta) = \cot\theta$.

\end{enumerate}
\end{question}

\section{Exercises}

\begin{exercise} \label{Qpfef34}
\begin{enumerate}
\item Find the $y$-coordinates of all points on the curve
\[
  y = f(\theta) = 10 - 4 \tan \theta
\]
where the tangent line is parallel to the line
\[
      8x - y = 12.
\]
Include a graph.
Start by defining an unkown like this:

Let 
\[
  (\theta_0 , f(\theta_0))
\]
be the coordinates of the point on the curve $y=f(\theta)$ where the tangent line is parallel ...

\item Repeat part (a) to find the $y$-coordinates of points on the same curve where the tangent line is parallel to the line
\[
   x - 8y = 12
\]
and then the line
\[
   8x + y = 12.
\]
Explain what's going on and why. Include a graph.
\end{enumerate}
\end{exercise}

\begin{exercise}
At some time in the afternoon a $40$-foot tall building casts an $80$-foot long shadow.

\begin{onlineOnly}
    \begin{center}
\desmos{5qqd2t4eny}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/5qqd2t4eny}{151: Building Shadow 7}

\begin{enumerate}
\item Use the worksheet above to approximate the scaling factor that converts a small change in the inclination angle of the sun (in radians) to an accurate approximation to the change in the length of the shadow. Include units. The tick marks on the protractor are spaced at equal intervals of $0.1$ radian.

\item Use calculus to find the exact scaling factor. Include units.

\item Find the length of the shadow when the scaling factor that converts a small change in the sun's inclination to an accurate approximation in the shadow's length is $-500$ ft/radian.
\end{enumerate}


\end{exercise}


\end{document}
