\documentclass{ximera}
\title{Differentiating Trigonometric Functions using Limits}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Differentiating trig functions from the definition of the derivative.
\end{abstract}
\maketitle



\section{The Motion of a Balloon}

\begin{example} \label{ELdebar3}

Play the slider in Line 1 of the worksheet below to watch the motion of a balloon. The balloon leaves behind tracks at equal intervals of time (see Line 4). Drag the slider $n$ in Line 2 to change the time interval.

\begin{onlineOnly}
    \begin{center}
\desmos{xsk95weesg}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/xsk95weesg}{151: Motion Cosine Derivative}

\begin{enumerate}
\item Sketch by hand a graph of the function
\[
    h  = f(t) \, , \, 0\leq t \leq 20
\]
that expresses the balloon's height (in feet) in terms of the number of seconds since the motion began. Label the axes with the appropriate variable names and units.

\item Activate the folder \emph{Graph} in Line 4 and play the motion again. How did you do?

\item Describe how the balloon's rate of ascent varies over the time of its motion. 

\item Are there time intervals during which the balloon's rate of ascent appears to be constant? If so, approximate these time intervals?

\item Use the motion to approximate the balloon's minimum and maximum rates of ascent. 

\item Sketch by hand a graph of the function
\[
     r = g(t) = \frac{dh}{dt} = \frac{d}{dt}\left( f(t) \right)
\]
that expresses the rate of ascent in terms of the number of seconds since the motion began. Label the axes with the appropriate variable names and units.
\end{enumerate}
\end{example}


\section{An Important Limit}
Computing the derivatives of the trigonometric functions hinges on one very important limit,
\[
 \lim_{\theta\to 0} \frac{\sin\theta}{\theta}.
\]

\begin{example} \label{ExMf2121}

For the worksheet below, do the following.
\begin{enumerate}
\item Express the $y$-coordinate of $P$ in terms of $a$ and $\theta$.
\[
    \answer{a\sin\theta}
\]

\item Express the length of arc $AP$ in terms of $a$ and $\theta$.
\[
    \answer{a\theta} 
\]

\item Drag the slider $u$ (another name for $\theta$ in Line 1 toward zero. What do you notice about the ratio $\sin\theta/\theta$ as you do this?
\begin{freeResponse}
\end{freeResponse}
\end{enumerate}

\begin{onlineOnly}
    \begin{center}
\desmos{2ocbmfdt5r}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/2ocbmfdt5r}{151: Sin theta over theta 0}


\end{example}


\begin{example}  \label{Ex8f7fd3lgzmq}
\begin{onlineOnly}
    \begin{center}
\desmos{xv3rn3tyxx}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/xv3rn3tyxx}{151: Sin theta over theta Limit}

\begin{enumerate}
\item Find an expression for the function
\[
      r = m(\theta)
\]
that gives the average rate of change of the function
\[
     f(\theta) = \sin \theta
\]
over the interval $[0,\theta]$. Include the domain.

\item Drag the slider $u$ in Line 6 above to sketch by hand a graph of the function $m$. Label the axes with the appropriate varaible names and units. Then activat the folder \emph{Average rate of change function} in Line 1 to see how you did. Zoom out to see more of the graph.

\item Approximate the limit
\[
     \lim_{\theta \to 0} \frac{\sin\theta}{\theta}
\]
in two ways:

\begin{enumerate}
\item Visually, by zooming in on the origin in the worksheet above.

\item Numerically, with a table of average rates of change. Do this with a calculator and then open the \emph{average rates of change} folder in Line 4 as a check.
\end{enumerate}

\item What's your guess?
\[
   \lim_{\theta \to 0} \frac{\sin\theta}{\theta} = \answer{1} .
\]

\end{enumerate}
\end{example}

\section{Variations on the Limit}

\begin{example}  \label{Exgzmq}
The purpose of this example is to evaluate the limit
\[
  \lim_{\theta\to 0} \frac{\sin (2\theta)}{\theta}.
\]

\begin{onlineOnly}
    \begin{center}
\desmos{u1czzdozr7}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/u1czzdozr7}{151: Sin 2theta over theta}

\begin{enumerate}
\item Interpret the expressions $a \sin 2\theta$ and $a \theta$ geometrically in the worksheet above.
\begin{freeResponse}
\end{freeResponse}

\item Drag the slider $u$ (another name for $\theta$) in Line 2 to guess the value of the limit
\[
   \lim_{\theta\to 0} \frac{\sin (2\theta)}{\theta}.
\]
\begin{freeResponse}
\end{freeResponse}

\item To compute the limit we'll make the substitution $u = 2\theta$. Then
\[
   \theta = \answer{u/2}
\]
and as $\theta\to 0$, 
\[
  u = 2\theta \to \answer{0}.
\]
So
\begin{align*}
\lim_{\theta\to 0} \frac{\sin (2\theta)}{\theta} &= \lim_{u\to \answer{0}} \frac{\answer{\sin u}}{\answer{u/2}} \\
                               &= \answer{2} \lim_{u \to 0} \frac{\answer{\sin u}}{\answer{u}} \\
                               &= 2 (\answer{1}) \\
                               &= \answer{2} .
\end{align*}

\end{enumerate}
\end{example}

\section{Sine Proof 1}

\begin{example}  \label{ExMV21er}
Here's one way to show
\[
   \frac{d}{d\theta}\left(  \sin\theta \right) = \cos\theta.
\]
 It uses the identity
\[
 \sin(\alpha + \beta) = \sin\alpha \cos\beta + \cos \alpha \sin\beta.
\]


\begin{enumerate}

\item First use the result of Example 2 to evaluate the derivative
\[
   \frac{d}{d\theta}\left(  \sin\theta \right) \Big|_{\theta=0} .
\]

\item Use the graph of the function $y=\cos\theta$ to evaluate the derivative
\[
  \frac{d}{d\theta}\left(  \cos\theta \right) \Big|_{\theta=0} .
\]
Explain your reasoning with a picture.
\begin{freeResponse}
\end{freeResponse}


\item Explain why
\[
      \frac{d}{d\theta}\left(  \sin\theta \right) \Big|_{\theta = \theta_0} = \frac{d}{d\theta}\left(  \sin(\theta + \theta_0)\right) \Big|_{\theta = 0}.
\]
Do this by graphing the two functions.
\begin{freeResponse}
\end{freeResponse}

\item Then use parts (a)-(c).
\begin{align*}
 \frac{d}{d\theta}\left(  \sin\theta \right) \Big|_{\theta = \theta_0} &= \frac{d}{d\theta}\left(  \sin(\theta + \theta_0)\right) \Big|_{\theta = 0} \\
                    &= \frac{d}{d\theta} \left( \sin\theta\cos\theta_0 + \cos\theta \sin\theta_0  \right)\Big|_{\theta = \theta_0} \\
                    &= \frac{d}{d\theta} \left( \sin\theta\cos\theta_0 \right)\Big|_{\theta = \theta_0} + \frac{d}{d\theta} \left( \cos\theta\sin\theta_0 \right)\Big|_{\theta = \theta_0}  \\
                    &= \left( \cos\theta_0 \right)  \frac{d}{d\theta} \left( \sin\theta \right)\Big|_{\theta = \theta_0} + (\sin\theta_0) \frac{d}{d\theta}\left( \cos\theta \right)\Big|_{\theta = \theta_0} \\
                    &= (\cos\theta_0)(\answer{1}) + (\sin\theta_0)(\answer{0})  \\
                    & = \answer{\cos} \theta_0 .
\end{align*}

\end{enumerate}
\end{example}



\section{Sine Proof 2}

\begin{example} \label{JJJD3}

Here's another way to show
\[
   \frac{d}{d\theta}\left(  \sin\theta \right) = \cos\theta ,
\]
this time with limits. It uses the identity
\[
 \sin\theta_2 - \sin \theta_1 = 2\sin\left( \frac{1}{2}\left( \theta_2 - \theta_1 \right)\right) \cos \left( \frac{1}{2}\left( \theta_2 +\theta_1 \right)\right) .
\]

\begin{enumerate}

\item Use the picture below to prove the above identity, at least when $0 < \theta_1 < \theta_2 < \pi/2$.

\begin{onlineOnly}
    \begin{center}
\desmos{yppjwpuhkz}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/yppjwpuhkz}{151: Difference of Sines}

\item Use the identity to find an expression for
\[
       \frac{d}{d\theta}\left(  \sin\theta \right)\Big|_{\theta = \theta_1} = \lim_{\theta_2\to\theta_1} \frac{\sin\theta_2 - \sin\theta_1}{\theta_2 - \theta_1}.
\]
\begin{freeResponse}
\end{freeResponse}

\end{enumerate}
\end{example}


\section{The Derivative of the Cosine Function}

\begin{example} \label{EppfemMvV}

\begin{enumerate}

\item One of the functions below is the derivative of the other. Which is which?
\begin{freeResponse}
\end{freeResponse}

\begin{onlineOnly}
    \begin{center}
\desmos{atxecntid1}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/atxecntid1}{151: Sine Cosine Derivative}

\item Drag the slider $k$ in Line 1 to find an expression for the derivative
\[
  \frac{d}{d\theta}\left(  \cos\theta \right).
\]
Explain your reasoning. Include graphs to help with your explanation.
\begin{freeResponse}
\end{freeResponse}

\end{enumerate}

\end{example}

\section{Exercises}

\begin{exercise} \label{Pf3mma}
\begin{enumerate}

\item Find a simplified equation of the tangent line to the curve
\[
      y = f(\theta) =  10 - 8\sin\theta
\]
at the point $P$ withcoordinates
\[
      (\theta_0 , f(\theta_0)),
\]
where
\[
    \theta_0 = \pi - \arcsin(3/4) .
\]
Do this \emph{without} a calcluator.

\item Check your work in the desmos worksheet below by entering the coordinates of $P$ in Line 3 and an equation of the tangent line in Line 4.

\begin{onlineOnly}
    \begin{center}
\desmos{sizljkxbth}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/sizljkxbth}{151: Tangent to Graph Sine 1}
\end{enumerate}
\end{exercise}

\begin{exercise} \label{ExLm383}
\begin{enumerate}
\item Drag the slider in Line 1 in the worksheet below to approximate the $y$-coordinate(s) of all points on the curve
\[
    y = f(\theta) = 10 - 12\cos \theta
\] 
where the tangent line is parallel to the line
\[
  3x + y = 15.
\]

\item Find simplified expressions for the exact $y$-coordinates without using a calculator. 

Start your solution as follows:

Let $(\theta_0,f(\theta_0))$ be the coordinates of a point on the curve $y=f(\theta)$ where the tangent line is parallel to the line 
\[
      3x + y = 15.
\]
Then ...


\begin{onlineOnly}
    \begin{center}
\desmos{8zwsudnvn0}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/8zwsudnvn0}{151: Tangent to Graph Cosine 1}


\end{enumerate}
\end{exercise}


\begin{exercise}  \label{Ex983vbvx}
\begin{enumerate}
\item Use the graph below to approximate the minimum and maximum outputs of the function
\[
      f(\theta) = 3\sin\theta - 5\cos\theta .
\] 

\begin{onlineOnly}
    \begin{center}
\desmos{msxpibv77r}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/msxpibv77r}{151: Graph Sine Plus Cosine 1}

\item Use calculus, \emph{not a trigonometric identity}, to find the exact minimum and maximum outputs.

Start your solution as follows:

Let $P$ with coordinates $(\theta_0 , f(\theta_0))$ be a point on the curve
\[
      y = f(\theta) =  3\sin\theta - 5\cos\theta .
\] 
where the output is a minimum or a maximum. Then the tangent line to the curve at $P$ is horizontal and
\[
     \frac{d}{d\theta} \left( 3\sin\theta - 5\cos\theta \right)\Big|_{\theta=\theta_0} = \answer{0}...
\]

\item Enter equations of the two horizontal lines though the turning points in Lines 1 and 2 of the worksheet to check your work.
\end{enumerate}

\end{exercise}


\begin{exercise} \label{ExKmd331}
A ferris wheel has a radius of $40$ feet and its center is $50$ feet above the ground. You ride the wheel for one revolution and then get off.

\begin{enumerate}
\item Find a function
\[
    h = f(\theta) \, , \, 0 \leq \theta \leq 2\pi ,
\]
that expresses your height above the ground (in feet) in terms of the rotation angle of the wheel (measured in radians with respect to the downward-pointing vertical.)

\item The wheel comes to a stop for the first time at some rotation angle $\theta = \theta_0$ when you are $80$ above the ground for the second time.

\begin{enumerate}
\item Use numerical methods to approximate the scaling factor that converts a small change 
\[
 \Delta \theta = \theta - \theta_0
\]
in the rotation angle to an accurate approximation to the corresponding change
\[
  \Delta h = f(\theta_0 + \Delta \theta) - f(\theta_0)
\]
in your height above the ground.

\item Use the derivative, not vectors, to find the exact value of the scaling factor.

\item What are the units of the scaling factor?

\item Interpret the scaling factor geometrically.

\item Suppose the wheel turns through the angle $\Delta \theta = 0.05$ radians and stops again. Use the previous result to approximate your new height.

\end{enumerate}

\end{enumerate}

\end{exercise}


\end{document}