\documentclass{ximera}
\title{Measuring Distances on the Earth}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Shortest paths on the earth.
\end{abstract}
\maketitle


\section{Along a Circle of Latitude}

\begin{question} \label{Q34dgbnhhtrg}

Pick points $A$ and $B$ at the same latiitude $\phi$ on the earth's surface and let $\theta$ (measured in radians) be the difference in their longitudes, with $0<\theta \leq \pi$.

\begin{onlineOnly}
   \begin{center}
\desmosThreeD{ajoeay29il}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/ajoeay29il}{151: Distances on Earth}

The purpose of this problem is to compare two distances in traveling from $A$ to $B$, one along their common circle of latitude, the other along the great circle through $A$ and $B$. We'll assume the earth is a sphere of radius $R$ miles.

To compute each distance, we need to see inside the earth. You can do this by deactivating the \emph{Sphere} folder in Line 3.

\begin{enumerate}
\item First the distance along the circle of latitude.

\begin{enumerate}
\item Start by expressing the radius of the circle of latitude in terms of the latitude $\phi$ and the radius of earth $R$. Work in general, not with the specific values in the worksheet above.

\item Then express the distance between $A$ and $B$ along the circle of latitude in terms of $\phi$, $R$, and $\theta$.
\end{enumerate}

\item Next we'll compute the distance along the great circle through $A$ and $B$. The center of this circle coincides with the center of the sphere.

\begin{enumerate}
\item Try to do this on your own.
\end{enumerate}

\item Compare these two distances betweeen San Francisco (latitude $38^\circ$N, longtitude $122^\circ$W) and Washington DC  (latitude $38^\circ$N, longtitude $77^\circ$W). Take the radius of the earth to be $3960$ miles.

\item Find two other locations at approximately the same latitude and compare the two distances.

\end{enumerate}

\end{question}


\end{document}

