\documentclass{ximera}
\title{A Balloon's Rate of Ascent}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Graphing an average rate of change function and using limits.
\end{abstract}
\maketitle




\begin{question}  \label{Q4gdfeft4tr4}
The function
\[
      h = f(t)  \, , \, 0\leq t \leq 2.2 ,                     %= -2t^3+7t^2-8t+8 \, , \, 0\leq t \leq 2.2 ,
\]
expresses the height of a balloon (in hundreds of feet) in terms of the number of minutes past noon.

The graph of the function $h=f(t)$ is shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{yd4xm6x6ub}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/yd4xm6x6ub}{151: Balloon}

\begin{enumerate}

\item Let $b\in [0,2.2]$ be a constant. Interpret the meaning of the function
\[
   m(v) =  \frac{f(v)-f(b)}{v-b}
\]
in the context of this scenario. Include units. State the domain of the function as well.

\item Use the graph of the function $h=f(t)$ above and the slider $u$ to to sketch a rough graph of the function $r=m(v)$ when $b=2$. Label the axes with the appropriate variable names and units.

\item Activate the folder (avg. rate of change function) in Line 11 to check your graph of the function $r=m(v)$. 

\item Click on the appopriate points of the curve $r=m(v)$ to find the balloon's average rate of ascent between 

\begin{enumerate}
\item 12:00pm and 12:02pm

\item 12:01pm and 12:02pm
\end{enumerate}

\item Zoom in close enough near the missing point on the curve $r=m(v)$ to approximate the balloon's rate of ascent at 12:02pm.

\item Open the Table in Line 1 by clicking the \emph{right arrow} at the left of the line. Use it to guess the instantaneous rate of ascent at 12:02pm.

\item Now suppose
\[
   f(t) = -2t^3+7t^2-8t+8 \, , \ 0\leq t \leq 2.2.
\]

\begin{enumerate}
\item Find a simplified expression for the function $r=m(v)$. Include the appropriate domain.

\item Use your simplified expression and limits to compute the balloon's rate of ascent at 12:02pm.

\item Assuming a constant rate of ascent, find the best possible approximation to the change
\[
     \Delta h = f(t) - f(2)
\]
in the balloon's height (in hundreds of feet) over the time interval
\[
  \Delta t = t - 2
\]
from 12:02pm to time $t$ minutes past noon. 

\item Use the result of part (iii) to approximate when the balloon is $380$ feet high. Explain your reasoning.

\item Use the graph of the function $h=f(t)$ to determine if your estimate in part (iii) is before or after the actual time. Explain your reasoning.



\end{enumerate}

\end{enumerate}
\end{question}



\end{document}