\documentclass{ximera}
\title{Substitution and Parametrically-defined Curves}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Substitution and parameterized curves
\end{abstract}
\maketitle

\section{The Basics}

\begin{question} \label{QkJDFE33d}
Which of the following definite integrals are equal to $\int_2^6 f(x)\, dx$? Avoid using a $u$-substitution. Instead think geometrically by describing the transformations that take the graph of $y=f(x)$ to the graph of each integrand. Assume $f$ is continuous.

\begin{enumerate}
\item $\int_4^{12} 2f(2x)\, dx$

\item $\int_4^{12} (1/2) f(x/2)\, dx$

\item $\int_1^{3} (1/2)f(2x)\, dx$

\item $\int_1^{3} 2f(2x)\, dx$
\end{enumerate}
\end{question}


\begin{question} \label{Q99frr3fds}
The function 
\[
     r = f(t) \, , \, 0\leq t \leq 12,
\]
expresses a balloon's rate of ascent (in ft/min) in terms of the number of minutes past noon. Its graph is shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{yp3awkyck8}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/yp3awkyck8}{152:Balloon 11}

\begin{enumerate}
\item When is the balloon the same height as it is at 12:02pm? Explain.

\item Use the graph to approximate the time(s) when the balloon is $20$ feet higher than it was at 12:02pm.

\item Write an equation whose solution(s) give the time(s) when the balloon is $20$ feet higher than it was at 12:02pm.

\item Suppose
\[
      f(t)=80\sin\left( \frac{\pi}{6}t \right) .
\]
Use a $u$-substitution and algebra to find exact solution(s) to your equation from part (c). Then use a calculator to approximate the clock time(s) to the nearest second.
\end{enumerate}
\end{question}



\begin{question} \label{QKdfrerr3}
Between noon and 8pm the growth rate of a population of bacteria decreases exponentially. At noon the population increases at the rate of $700,000$ bac/hr and at 3pm the population increases at the rate of $400,000$ bac/hr. 

The population is $2$ million at noon.

\begin{enumerate}

\item Sketch by hand a rough graph of the function
\[
    r = f(t) \, , \, 0\leq t \leq 8,
\]
that expresses the growth rate (in hundreds of thousand of bacteria/hr) in terms of the number of hours past noon.

\item Find an expression for the function in part (a). Do not use $e$.

\item Use part (a) to sketch by hand a graph of the function 
\[
 P=g(t) \, , \, 0\leq t \leq 8, 
\]
that expresses the population in terms of the number of hours past noon.

\item Find an expression for the function in part (c).

\end{enumerate}
\end{question}

\begin{question}\label{Q7yyweff}
Identify for which of the following definite integrals a $u$-substitution would work. Make the substitution for these and evaluate the integral. Do this by changing the lower and upper bounds. Do \emph{not} substitute back to the original variable of integration.

\begin{enumerate}

\item $\int_0^{\sqrt{\pi}} \theta \sin(\theta^2)\, d\theta$

\item $\int_0^{\pi} \theta^2 \sin(\theta)\, d\theta$

\item $\int_0^{\pi/2} \sin\theta \cos^3\theta \, d\theta$

\item $\int_0^{\pi/4} \tan\theta \sec^2\theta \, d\theta$

\item $\int_0^{\pi/4} \tan\theta \sec^3\theta \, d\theta$

\item $\int_0^2 x^2 (1+x^2)^2 \, dx$

\item $\int_0^2 x^2 (1+x^3)^2 \, dx$

\item $\int_1^2 \frac{\cos \left( \sqrt{x} \right)}{\sqrt{x}}\, dx$

\item $\int_1^2 \sqrt{x} \cos \left( \sqrt{x} \right)\, dx$

\item $\int_0^1 \frac{e^x}{(1+e^x)^2} \ , dx$

\item $\int_0^1 \frac{e^x}{1+e^x} \ , dx$

\item $\int_0^1 \frac{e^x}{1+e^{2x}} \ , dx$

\item $\int_2^3  \frac{x}{\ln x}\, dx$

\item $\int_2^3  \frac{\ln x}{x}\, dx$


\end{enumerate}

\end{question}


\section{A Geometric Interpretation}
To understand what $u$-substitution does geometrically, it helps to think about making a \emph{reverse} $u$-substitution. This means starting with a simple integral and making it more complicated with a substitution. Here's an example.

\begin{example} \label{EKEMrerMMEF}
We'll start with the integral
\[
      I = \int_0^\pi \sin u \, du
\]
and make the substitution 
\[
     u = t^2 .
\]
This substitution really reparameterizes the curve
\[
   y = f(u) = \sin u \, , \, 0\leq u \leq \pi
\] 
from the standard parameterization
\[
   (x,y) = (u,\cos u) \, , \, 0\leq u \leq \pi
\]
to the new parameterization
\[
   (x,y) = (t^2,\cos (t^2) \, , \, 0\leq t \leq \sqrt{\pi}.
\]

The substitution maps the rectangle with height $y=\sin u$ and width $du$ to the rectangle with the same height $y=\sin(t^2)$ and width
\[
   du = 2t \,dt .
\]
The effect is to transform a partition of the interval of integration, originally $u\in [0,\pi]$ with equal subintervals into a partition of the new interval of integration $t\in [0,\sqrt{2\pi}]$ with \emph{unequal} subintervals $du=2t\, dt$. You can see this transformation by dragging the slider $w$ in Line 2 of the worksheeet below from $w=0$ to $w=1$.



\begin{onlineOnly}
    \begin{center}
\desmos{toiwzzqkva}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/toiwzzqkva}{152: Sub 2}

\end{example}

The next example puts this idea in more a more meaningul context.

\begin{example} \label{Ex:KDfdfer3rER}

The function
\[
    r = f(s) \, , \, 0 \leq s \leq 240,
\]
expresses the rate at which a car burns gas (in gal/mile) in terms of the trip odometer reading (in miles) during a $240$-mile trip.

The car has $14$ gallons of gas at odometer reading $100$ miles.

The function 
\[
      s = g(t) \, , \, 0\leq t \leq 4 ,
\]
expresses the car's speed (in miles/hour) in terms of the number of hours since the start of the trip.

\begin{enumerate}
\item Find a function 
\[
      G = w(s) \, , \, 0 \leq s \leq 240,
\]
that expresses the number of gallons of gas in the car in terms of the trip odometer reading.

\item Make a substitution in your expression from part (a) to find a function 
\[
  G = h(t) \, , \, 0 \leq s \leq 240,
\]
that expresses the number of gallons of gas in the car in terms of the number of hours since the start of the trip.

\item Your function for part (b) should be
\[
     G = h(t) = 14 - \int_0^t f(g(u))g^\prime(u)  \, du .
\]

Interpret the integral in this expression in two different ways.

\begin{enumerate}
\item By writing it as
\[
   \int_0^t f(g(u))(  g^\prime(u)  \, du )
\]
and describing the meaning of the product $g^\prime(u)\, du$.

\item By keeping its original form and describing the meaning of the product $f(g(u))g^\prime(u)$.

\end{enumerate}

\item Drag the slider $w$ in Line 2 of the worksheet below to change the odometer reading function $g$. Describe what happens.

\end{enumerate}

\begin{onlineOnly}
    \begin{center}
\desmos{gsdzd0lbva}{450}{600}  
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/gsdzd0lbva}{Substitution and Gas Comsumption 2}

\end{example}



\end{document}

