\documentclass{ximera}
\title{The Chain Rule and the Price of a Stock}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
A stock problem with the chain rule.
\end{abstract}
\maketitle


\begin{question} \label{QLfel33mz}
The function
\[
     P = f(t) = 64 - 20\cos\left(  \frac{t}{4} \right) \, , \, 8\leq t \leq 16,
\]
expresses the price of a stock (in dollars/share) in terms of the number of hours past 9am.

\begin{enumerate}

\item What are the units of the constant $40$ in the function above? The constant $20$? The constant $1/4$? Explain your reasoning.

\item Sketch by hand a reasonably accurate graph of the function. Label the axes with the appropriate variable names and units. Include scales on both axes. Label the coordinates of any turning points.

\item Find an expression for the derivative $dP/dt$. Do this by making an explicit $u$-substitution and showing each step of the chain rule. But do \emph{not} use $u$. Use a more meaningful variable instead.

\item What are the units of the derivative $dP/dt$? Interpret its meaning.

\item Find the stock price when it is decreasing at the rate of ($\$3$/share)/hr.

\item Find a function 
\[
   P = g(r)
\]
that expresses the stock price (in dollars/share) in terms of the rate of change in the price (measured in (dollars/share)/hr). Simplify the expression for your function.

\item Sketch by hand a graph of the function $P=g(r)$. Label the axes with the appropriate variable names and units. 

\item Evaluate the derivative 
\[
  \frac{dP}{dr}\Big|_{r=-3}.
\] 
Include units and interpret its meaning as a scaling factor. Include an example of how the scaling factor acts on a \emph{specific} small change.

%\item Is the stock price increasing or decreasing the second time the stock sells for $\$45$/share? At what rate?

\item Is the number of shares you can buy with $\$1000$ increasing or decreasing when the price is decreasing at the rate of ($\$4$/share)/hour? 

\begin{enumerate}

\item Use numerical methods, no calculus, to approximate the rate of change (with respect to time) in the number of shares you can buy with $\$1,000$ when the price is decreasing at the rate of ($\$4$/share)/hour. No calculus.

\item Use calculus to compute the rate in part (i). Quotient rule not allowed.

\end{enumerate}

\end{enumerate}

\end{question}

\end{document}

